### 2015-09-29 manuscript.Rnw
finish studies example, fix citations

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies. In parallel, new phylogenetic comparative methods
have been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \cite{Pennell2013}. The development of these methods has
contributed to the need for a greater availability of high-quality phylogenetic
information across life. However, phylogenetic information is scattered, often
only available as image files within publications, and the variety of file
formats used to store phylogenetic data makes it difficult for researchers to
access, synthesize, and integrate this information into their own research
(\cite{Stoltzfus2012}, but see \cite{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and conservation, the
resources OTL provides could be used for education and outreach (e.g.,
illustrating course material, develop outreach activities to explore
relationships among species).

@@

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL's APIs. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
structure (objects of class \texttt{phylo}), or can be written to files in the
Newick, NEXUS \citep{Maddison1997} or NeXML \citep{Vos2012} file formats. This
allows researchers to use these trees either directly with other R packages, or
to be imported in other programs that can read in phylogenetic tree files.

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by the API. However, resources and methods are being developed to
add branch lengths information to these topological trees (e.g.,
\cite{Ksepka2015}) or use topological trees to identify phylogenetically
equivalent species to increase overlap between chronograms and species trait
data \cite{Pennell2015}. Without branch lengths, these trees remain useful to
illustrate relationships among species, or to map a trait on a phylogeny for
instance.

@@

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and also identify the trees that
are being used to assemble the synthetic tree. These trees provide a useful
resource to reproduce or expand on an already published analysis.

Criteria that can be used to search for studies or their associated trees are
available through the output of the function
\texttt{studies\_properties}. Typically, users will want to search for studies
or trees based on a taxon but other criteria such as the title of the
publication. Here we demonstrate how to look for and retrieve a tree for study
focusing on cats, Felidae (Figure~\ref{fig:plot_cats}).

<<plot_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
                                    value = "felidae")
cat_studies
cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
                           tree_id = list_trees(cat_studies)[[1]][1])
plot(cat_tree)
@

\subsection{Exploring the taxonomy}
\label{sec:exploring-taxonomy}

\subsection{How does rotl fit into the R package ecosystem?}
\label{sec:how-does-rotl-fit}

>>>>

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies. In parallel, new phylogenetic comparative methods
have been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \citep{Pennell2013}. The development of these methods has
contributed to the need for a greater availability of high-quality phylogenetic
information across life. However, phylogenetic information is scattered, often
only available as image files within publications, and the variety of file
formats used to store phylogenetic data makes it difficult for researchers to
access, synthesize, and integrate this information into their own research
(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and conservation, the
resources OTL provides could be used for education and outreach (e.g.,
illustrating course material, develop outreach activities to explore
relationships among species).

@@

\texttt{rotl} gives users access to all endpoints provided by version 2 of the
OTL's APIs. Phylogenetic trees served by the API can be imported directly into
R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
structure (objects of class \texttt{phylo}), or can be written to files in the
Newick, NEXUS \citep{Maddison1997} or NeXML \citep{Vos2012} file formats. This
allows researchers to use these trees either directly with other R packages, or
to be imported in other programs that can read in phylogenetic tree files.

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by the API. However, resources and methods are being developed to
add branch lengths information to these topological trees
\citep[e.g.,][]{Ksepka2015} or use topological trees to identify
phylogenetically equivalent species to increase overlap between chronograms and
species trait data \citep{Pennell2015}. Without branch lengths, these trees
remain useful to illustrate relationships among species, or to map a trait on a
phylogeny for instance.

@@

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and identify the trees that
contribute to the synthetic tree. These trees provide a useful resource to
reproduce or expand on an already published analysis.

Criteria that can be used to search for studies or their associated trees are
available through the output of the function
\texttt{studies\_properties}. Typically, users will want to search for studies
or trees based on taxon names (or their Open Taxonomy identifiers), but other
criteria such as the title of the publication can be used. Here we demonstrate
how to look for and retrieve a tree for study focusing on the family Felidae
(Figure~\ref{fig:plot_cats}).

<<get_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
                                    value = "felidae")
cat_studies
@

Currently only one study focused on this family is available from OTL, and a
single tree is associated with it. We can then retrieve the identifiers and pass
them to the function \texttt{get\_study\_tree} to have the tree in memory:

<<get_cat_tree, results='hide'>>=
cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
                           tree_id = cat_studies[["tree_ids"]][1])
cat_tree
@

<<print_cat_tree, echo=FALSE>>=
print(cat_tree, printlen = 4)
@

When more than one tree is available for a given study, the function
\texttt{list\_trees} returns a list to retrieve the tree identifiers for each
study easily. Alternatively, the function \texttt{get\_study} returns all the
trees associated with a particular study. Additional details about the study can
be obtained using the function \texttt{get\_study\_meta}.

\subsection{How does \texttt{rotl} fit into the R package ecosystem?}
\label{sec:how-does-rotl-fit}
